---
source: Rmd
title: "Experimental design"
teaching: XX
exercises: XX
questions:
- "How do we design experiments optimally?"
- "How do we interpret a given design?"
objectives:
- "Explain the formula notation and design matrices."
- "Explore different designs and learn how to interpret coefficients."
keypoints:
- "Key point 1"
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, purl=FALSE, message=FALSE}
if (!file.exists("data/GSE96870_se.rds")) {
    dir.create("data, showWarnings = FALSE")
    download.file(
        url = "https://github.com/Bioconductor/bioconductor-teaching/blob/master/data/GSE96870/GSE96870_se.rds?raw=true", 
        destfile = "data/GSE96870_se.rds"
    )
}
```

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
options(width = 120)
```

```{r}
suppressPackageStartupMessages({
    library(SummarizedExperiment)
    library(ExploreModelMatrix)
    library(dplyr)
})
```

```{r}
se <- readRDS("data/GSE96870_se.rds")
meta <- as.data.frame(colData(se))
meta
```

```{r}
vd <- VisualizeDesign(sampleData = meta, 
                      designFormula = ~ tissue + time + sex)
vd$cooccurrenceplots
```

# Simple example - compare males and females, non-infected spinal cord

```{r}
meta_noninf_spc <- meta %>% filter(time == "Day0" & 
                                       tissue == "Spinalcord")
meta_noninf_spc
vd <- VisualizeDesign(sampleData = meta_noninf_spc, 
                      designFormula = ~ sex)
vd$designmatrix
vd$plotlist
```

> ## Challenge: Can you do it?
>
> Set up the design formula to compare the three time points (Day0, Day4, Day8) in the male spinal cord samples, and visualize it using `ExploreModelMatrix`. 
>
>
> > ## Solution
> >
> > ```{r, echo=TRUE}
> > meta_male_spc <- meta %>% filter(sex == "Male" & tissue == "Spinalcord")
> > meta_male_spc
> > vd <- VisualizeDesign(sampleData = meta_male_spc, designFormula = ~ time)
> > vd$designmatrix
> > vd$plotlist
> > ```
> {: .solution}
{: .challenge}

# Factorial design without interactions

```{r}
meta_noninf <- meta %>% filter(time == "Day0")
meta_noninf
vd <- VisualizeDesign(sampleData = meta_noninf, 
                      designFormula = ~ sex + tissue)
vd$designmatrix
vd$plotlist
```

# Factorial design with interactions

```{r}
meta_noninf <- meta %>% filter(time == "Day0")
meta_noninf
vd <- VisualizeDesign(sampleData = meta_noninf, 
                      designFormula = ~ sex * tissue)
vd$designmatrix
vd$plotlist
```

# Paired design

```{r}
meta_fem_day0 <- meta %>% filter(sex == "Female" & 
                                     time == "Day0")
meta_fem_day0
vd <- VisualizeDesign(sampleData = meta_fem_day0,
                      designFormula = ~ mouse + tissue)
vd$designmatrix
vd$plotlist
```

# Within- and between-subject comparisons

```{r}
meta_fem_day04 <- meta %>% 
    filter(sex == "Female" & 
               time %in% c("Day0", "Day4")) %>%
    droplevels()
meta_fem_day04
design <- model.matrix(~ mouse, data = meta_fem_day04)
design <- cbind(design, 
                Spc.Day0 = meta_fem_day04$tissue == "Spinalcord" & 
                    meta_fem_day04$time == "Day0",
                Spc.Day4 = meta_fem_day04$tissue == "Spinalcord" & 
                    meta_fem_day04$time == "Day4")
rownames(design) <- rownames(meta_fem_day04)
design
vd <- VisualizeDesign(sampleData = meta_fem_day04 %>%
                          select(time, tissue, mouse),
                      designFormula = NULL, 
                      designMatrix = design, flipCoordFitted = FALSE)
vd$designmatrix
vd$plotlist
```

