---
title: "Importing and annotating quantified data into R"
source: Rmd
teaching: XX
output:
  html_document:
    df_print: paged
exercises: XX
---

```{r, echo=FALSE, purl=FALSE, message=FALSE}
source("download_data.R")
```

::::::::::::::::::::::::::::::::::::::: objectives
-   Learn how to import the quantifications into a SummarizedExperiment object.
-   Learn how to add additional gene annotations to the object.
::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions
-   How can one import quantified gene expression data into an object suitable for downstream statistical analysis in R?
-   What types of gene identifiers are typically used, and how are mappings between them done?
::::::::::::::::::::::::::::::::::::::::::::::::::

## Load libraries

In this episode we will use some functions from add-on R packages. In order to use them, we need to load them from our `library`:

```{r}
suppressPackageStartupMessages({
    library(AnnotationDbi)
    library(org.Mm.eg.db)
    library(hgu95av2.db)
    library(SummarizedExperiment)
})
```

If you get any error messages about `there is no package called 'XXXX'` it means you have not installed the package/s yet for this version of R. See the bottom of the [Summary and Setup](https://carpentries-incubator.github.io/bioc-rnaseq/index.html) to install all the necessary packages for this workshop. If you have to install, remember to re-run the `library` commands above to load them. 

## Load data

In the last episode, we used R to download 4 files from the internet and saved them on our computer. But we do not have these files loaded into R yet so that we can work with them. The original experimental design in [Blackmore et al. 2017](https://pubmed.ncbi.nlm.nih.gov/28696309/) was fairly complex: 8-12 weeks old male and female mice, with two different genetic backgrounds (wild-type and D2D), were collected at Day 0 (before influenza infection), Day 4 and Day 8 after influenza infection. From each mouse, cerebellum and spinal cord tissues were taken for RNA-seq. There were originally 4 mice per 'Sex X Day x Genotype' group, but a few were lost along the way resulting in a total of 45 samples. For this workshop, we are going to simplify the analysis by only using the 22 cerebellum samples. Expression quantification was done using STAR to align to the mouse genome and then counting reads that map to genes. In addition to the counts per gene per sample, we also need information on which sample belongs to which Sex/Time point/Replicate. And for the genes, it is helpful to have extra information called annotation.
Let's read in the data files that we downloaded in the last episode and start to explore them:


### Counts

```{r}
counts <- read.csv("data/GSE96870_counts_cerebellum.csv", 
                   row.names = 1)
dim(counts)
# View(counts)
```

Genes are in rows and samples are in columns, so we have counts for 41,786 genes and 22 samples. The `View()` command has been comment out for the website, but running it will open a tab in RStudio that lets us look at the data and even sort the table by a particular column. However, the viewer cannot change the data inside the `counts` object, so we can only look, not permanently sort nor edit the entries. When finished, close the viewer using the X in the tab. Looks like the rownames are gene symbols and the column names are the GEO sample IDs, which are not very informative for telling us which sample is what.

### Sample annotations

Next read in the sample annotations. Because samples are in columns in the count matrix, we will name the object `coldata`:

```{r}
coldata <- read.csv("data/GSE96870_coldata_cerebellum.csv",
                    row.names = 1)
dim(coldata)
# View(coldata)
```

Now samples are in rows with the GEO sample IDs as the rownames, and we have 10 columns of information. The columns that are the most useful for this workshop are `geo_accession` (GEO sample IDs again), `sex` and `time`.

### Gene annotations
The counts only have gene symbols, which while short and somewhat recognizable to the human brain, are not always good absolute identifiers for exactly what gene was measured. For this we need additional gene annotations that were provided by the authors. The `count` and `coldata` files were in comma separated value (.csv) format, but we cannot use that for our gene annotation file because the descriptions can contain commas that would prevent a .csv file from being read in correctly. Instead the gene annotation file is in tab separated value (.tsv) format. Likewise, the descriptions can contain the single quote `'` (e.g., 5'), which by default R assumes indicates a character entry. So we have to use a more generic function `read.delim()` with extra arguments to specify that we have tab-separated data (`sep = "\t"`) with no quotes used (`quote = ""`). We also put in other arguments to specify that the first row contains our column names (`header = TRUE`), the gene symbols that should be our `row.names` are in the 5th column (`row.names = 5`), and that NCBI's species-specific gene ID (i.e., ENTREZID) should be read in as character data even though they look like numbers (`colClasses` argument). You can look up this details on available arguments by simply entering the function name starting with question mark. (e.g., `?read.delim`)

```{r}
rowranges <- read.delim("data/GSE96870_rowranges.tsv", 
                        sep = "\t", 
                        colClasses = c(ENTREZID = "character"),
                        header = TRUE, 
                        quote = "", 
                        row.names = 5)
dim(rowranges)
# View(rowranges)
```

For each of the 41,786 genes, we have the `seqnames` (e.g., chromosome number), 
`start` and `end` positions, `strand`, `ENTREZID`, gene product description 
(`product`) and the feature type (`gbkey`). These gene-level metadata are 
useful for the downstream analysis. For example, from the `gbkey` column, we
can check what types of genes and how many of them are in our dataset:

```{r}
table(rowranges$gbkey)
```

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge: Discuss the following points with your neighbor

- Have you used spreadsheets, in your research, courses,
  or at home?
- What kind of operations do you do in spreadsheets?
- Which ones do you think spreadsheets are good for?
- Have you accidentally done something in a spreadsheet program that made you
  frustrated or sad?
  
1. How are the 3 objects `counts`, `coldata` and `rowranges` related to each other in terms of their rows and columns?
2. If you only wanted to analyse the mRNA genes, what would you have to do keep just those (generally speaking, not exact codes)?
3. If you decided the first two samples were outliers, what would you have to do to remove those (generally speaking, not exact codes)?
  
::::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::: solution

1. In `counts`, the rows are genes just like the rows in `rowranges`. The columns in `counts` are the samples, but this corresponds to the rows in `coldata`. 
2. I would have to remember subset both the rows of `counts` and the rows of `rowranges` to just the mRNA genes.
3. I would have to remember to subset both the columns of `counts` but the rows of `coldata` to exclude the first two samples.

:::::::::::::::::::::::::::::::::::

You can see how keeping related information in separate objects could easily lead to mis-matches between our counts, gene annotations and sample annotations. This is why Bioconductor has created a specialized S4 class called a `SummarizedExperiment`. The details of a `SummarizedExperiment` were covered extensively at the end of the [Introduction to data analysis with R and Bioconductor](https://carpentries-incubator.github.io/bioc-intro/60-next-steps.html#next-steps) workshop. 
As a reminder, let's take a look at the figure below representing the anatomy of the `SummarizedExperiment` class:

```{r SE, echo=FALSE, out.width="80%"}
knitr::include_graphics("https://uclouvain-cbio.github.io/WSBIM1322/figs/SE.svg")
```

It is designed to hold any type of quantitative 'omics data (`assays`) along with linked sample annotations (`colData`) and feature annotations with (`rowRanges`) or without (`rowData`) chromosome, start and stop positions. Once these three tables are (correctly!) linked, subsetting either samples and/or features will correctly subset the `assay`, `colData` and `rowRanges`. Additionally, most Bioconductor packages are built around the same core data infrastructure so they will recognize and be able to manipulate `SummarizedExperiment` objects. Two of the most popular RNA-seq statistical analysis packages have their own extended S4 classes similar to a `SummarizedExperiment` with the additional slots for statistical results: [DESeq2](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#the-deseqdataset)'s `DESeqDataSet` and [edgeR](https://www.rdocumentation.org/packages/edgeR/versions/3.14.0/topics/DGEList-class)'s `DGEList`. No matter which one you end up using for statistical analysis, you can start by putting your data in a `SummarizedExperiment`. 

## Assemble SummarizedExperiment
We will create a `SummarizedExperiment` from these objects:

- The `count` object will be saved in `assays` slot    
- The `coldata` object with sample information will be stored in `colData` slot (_**sample metadata**_)    
- The `rowranges` object describing the genes will be stored in `rowRanges` slot (_**features metadata**_)     

Before we put them together, you ABSOLUTELY MUST MAKE SURE THE SAMPLES AND GENES ARE IN THE SAME ORDER! Even though we saw that `count` and `coldata` had the same number of samples and `count` and `rowranges` had the same number of genes, we never explicitly checked to see if they were in the same order. One quick way to check:


```{r}
all.equal(colnames(counts), rownames(coldata)) # samples
all.equal(rownames(counts), rownames(rowranges)) # genes

# If the first is not TRUE, you match up the samples/columns in counts
# with the samples/rows in coldata like this (which is fine to run
# even if the first was TRUE):

tempindex <- match(colnames(counts), rownames(coldata))
coldata <- coldata[tempindex, ]

all.equal(colnames(counts), rownames(coldata)) 
# Should now be TRUE

```

:::::::::::::::::::::::::::::::::::::::  challenge

If the features (i.e., genes) in the assay (e.g., `counts`) and the gene
annotation table (e.g., `rowranges`) are different, how can we fix them? 
Write the codes. 

:::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::: solution

```{r eval=FALSE}
tempindex <- match(rownames(counts), rownames(rowranges))
coldata <- coldata[tempindex, ]

all.equal(rownames(counts), rownames(rowranges)) 


```


:::::::::::::::::::::::::::::::::::




Once we have verified that samples and genes are in the same order, we can 
then create our `SummarizedExperiment` object.

```{r}
# One final check:
stopifnot(rownames(rowranges) == rownames(counts), # features
          rownames(coldata) == colnames(counts)) # samples

se <- SummarizedExperiment(
    assays = list(counts = as.matrix(counts)),
    rowRanges = as(rowranges, "GRanges"),
    colData = coldata
)
```


A brief recap of how to access the various data slots in a `SummarizedExperiment`:

```{r}
# Access the counts
head(assay(se))
dim(assay(se))

# Access the sample annotations
colData(se)
dim(colData(se))

# Access the gene annotations
head(rowData(se))
dim(rowData(se))
```



## Save SummarizedExperiment

This was a bit of code and time to create our `SummarizedExperiment` object. We will need to keep using it throughout the workshop, so it can be useful to save it as an actual single file on our computer to read it back in to R's memory if we have to shut down RStudio. To save an R-specific file we can use the `saveRDS()` function and later read it back into R using the `readRDS()` function. 

```{r}
saveRDS(se, "data/GSE96870_se.rds")
rm(se) # remove the object!
se <- readRDS("data/GSE96870_se.rds")
```


## Annotations
Depending on who generates your count data, you might not have a nice file of 
additional gene annotations. There may only be the count row names, which 
could be gene symbols or ENTREZIDs or another database's ID. Characteristics 
of gene annotations differ based on their annotation strategies and information 
sources. For example, RefSeq human gene models (i.e., Entrez from NCBI) are 
well supported and broadly used in various studies. The UCSC Known Genes 
dataset is based on protein data from Swiss-Prot/TrEMBL (UniProt) and the 
associated mRNA data from GenBank, and serves as a foundation for the UCSC 
Genome Browser. Ensembl genes contain both automated genome annotation and 
manual curation.

You can find more information in Bioconductor [Annotation Workshop](https://jmacdon.github.io/Bioc2022Anno/articles/AnnotationWorkshop.html)
material.

Bioconductor has many packages and functions that can help you to get additional annotation information for your genes. The available resources are covered in more detail in [Episode 6 Gene set enrichment analysis](https://carpentries-incubator.github.io/bioc-rnaseq/06-gene-set-analysis.html#gene-set-resources). 

Here, we will introduce one of the gene ID mapping functions, `mapIds`:
```
mapIds(annopkg, keys, column, keytype, ..., multiVals)
```

Where 
- *annopkg* is the annotation package        
- *keys* are the IDs that we **know**       
- *column* is the value we **want**    
- *keytype* is the type of key used    

```{r}
mapIds(org.Mm.eg.db, keys = "497097", column = "SYMBOL", keytype = "ENTREZID")
```

Different from the `select()` function, `mapIds()` function handles 1:many 
mapping between keys and columns through an additional argument, `multiVals`.
The below example demonstrate this functionality using the `hgu95av2.db` 
package, an Affymetrix Human Genome U95 Set annotation data.

```{r}
keys <- head(keys(hgu95av2.db, "ENTREZID"))
last <- function(x){x[[length(x)]]}

mapIds(hgu95av2.db, keys = keys, column = "ALIAS", keytype = "ENTREZID")
mapIds(hgu95av2.db, keys = keys, column = "ALIAS", keytype = "ENTREZID", multiVals = last)
mapIds(hgu95av2.db, keys = keys, column = "ALIAS", keytype = "ENTREZID", multiVals = "list")
```


:::::::::::::::::::::::::::::::::::::::::  callout

## Important!

Saving an object as an .rds file and reading it back in can be a convenient time saver for an extended analysis. However, there is no record in the object or .rds file of how it was made or manipulated. Your code to make the .rds file can become lost or corrupted, such that you cannot create the same object again from input data files. If this happens you should consider the .rds file as lost/corrupted as well. For reproducibility, the codes are more important than the actual .rds file.

::::::::::::::::::::::::::::::::::::::::::::::::::


## Session info

```{r}
sessionInfo()
```

::: keypoints
-   Depending on the gene expression quantification tool used, there are different ways (often distributed in Bioconductor packages) to read the output into a `SummarizedExperiment` or `DGEList` object for further processing in R.
-   Stable gene identifiers such as Ensembl or Entrez IDs should preferably be used as the main identifiers throughout an RNA-seq analysis, with gene symbols added for easier interpretation.
:::
