---
source: Rmd
title: Importing and annotating quantified data into R
teaching: XX
exercises: XX
---

```{r, echo=FALSE, purl=FALSE, message=FALSE}
source("download_data.R")
```

::::::::::::::::::::::::::::::::::::::: objectives

- Learn how to import the quantifications into a SummarizedExperiment object.
- Learn how to add additional gene annotations to the object.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- How do we get our data into R?

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("04-")
```

:::::::::::::::::::::::::::::::::::::::::  callout

### Contribute!

This episode is intended to show how we can assemble a SummarizedExperiment
starting from individual count, rowdata and coldata files. Moreover, we will
practice adding annotations for the genes, and discuss related concepts
and things to keep in mind (annotation sources, versions, 'helper' packages
like tximeta).


::::::::::::::::::::::::::::::::::::::::::::::::::

## Read the data

### Counts

```{r}
counts <- read.csv("data/GSE96870_counts_cerebellum.csv", 
                   row.names = 1)
```

### Sample annotations

```{r}
coldata <- read.csv("data/GSE96870_coldata_cerebellum.csv",
                    row.names = 1)
```

### Gene annotations

Need to be careful - the descriptions contain both commas and ' (e.g., 5')

```{r}
rowranges <- read.delim("data/GSE96870_rowranges.tsv", sep = "\t", 
                        colClasses = c(ENTREZID = "character"),
                        header = TRUE, quote = "", row.names = 5)
```

Mention other ways of getting annotations, and practice querying org package.
Important to use the right annotation source/version.

```{r}
suppressPackageStartupMessages({
    library(org.Mm.eg.db)
})
mapIds(org.Mm.eg.db, keys = "497097", column = "SYMBOL", keytype = "ENTREZID")
```

Check feature types

```{r}
table(rowranges$gbkey)
```

## Assemble SummarizedExperiment

```{r}
stopifnot(rownames(rowranges) == rownames(counts),
          rownames(coldata) == colnames(counts))

se <- SummarizedExperiment(
    assays = list(counts = as.matrix(counts)),
    rowRanges = as(rowranges, "GRanges"),
    colData = coldata
)
```

## Save SummarizedExperiment

```{r}
saveRDS(se, "data/GSE96870_se.rds")
```

## Session info

```{r}
sessionInfo()
```

:::::::::::::::::::::::::::::::::::::::: keypoints

- Key point 1

::::::::::::::::::::::::::::::::::::::::::::::::::


