---
source: Rmd
title: Extra exploration of design matrices
teaching: 30
exercises: 30
editor_options:
  chunk_output_type: console
---

```{r, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
    warning = FALSE
)
```


```{r, echo=FALSE, purl=FALSE, message=FALSE}
source("download_data.R")
```

::::::::::::::::::::::::::::::::::::::: objectives

- Explain the formula notation and design matrices.
- Explore different designs and learn how to interpret coefficients.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- How can one translate biological questions and comparisons to statistical terms suitable for use with RNA-seq analysis packages?

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, include=FALSE}
```

## Loading required packages and reading data

We start by loading a few packages that will be needed in this episode. 
In particular, the [ExploreModelMatrix](https://bioconductor.org/packages/ExploreModelMatrix/) package provides resources for exploring design matrices in a graphical fashion, for easier interpretation. 

```{r}
suppressPackageStartupMessages({
    library(SummarizedExperiment)
    library(ExploreModelMatrix)
    library(dplyr)
})
```

Next, we read the metadata table for our data set. 
As seen in previous episodes, the metadata contains information about the age, sex, infection status, time of measurement and tissue of the collected samples. 
Note that Day0 always corresponds to non-infected samples, and that infected samples are collected on days 4 and 8.
Moreover, all mice have the same age (8 weeks). 
Hence, in the first part of this episode we consider only the sex, tissue and time variables further. 

```{r}
meta <- read.csv("data/GSE96870_coldata_all.csv", row.names = 1)
meta
table(meta$time, meta$infection)
table(meta$age)
```

We can start by visualizing the number of observations for each combination of the three predictor variables. 

```{r}
vd <- VisualizeDesign(sampleData = meta, 
                      designFormula = ~ tissue + time + sex)
vd$cooccurrenceplots
```

:::::::::::::::::::::::::::::::::::::::  challenge

### Challenge

Based on this visualization, would you say that the data set is balanced, or are there combinations of predictor variables that are severely over- or underrepresented?

::::::::::::::::::::::::::::::::::::::::::::::::::


## Compare males and females, non-infected spinal cord

Next, we will set up our first design matrix. 
Here, we will focus on the uninfected (Day0) spinal cord samples, and our aim is to compare the male and female mice. 
Thus, we first subset the metadata to only the samples of interest, and next set up and visualize the design matrix with a single predictor variable (sex). 
By defining the design formula as `~ sex`, we tell R to include an intercept in the design. 
This intercept will represent the 'baseline' level of the predictor variable, which in this case is selected to be the Female mice. 
If not explicitly specified, R will order the values of the predictor in alphabetical order and select the first one as the reference or baseline level. 

```{r}
## Subset metadata
meta_noninf_spc <- meta %>% filter(time == "Day0" & 
                                       tissue == "Spinalcord")
meta_noninf_spc

## Use ExploreModelMatrix to create a design matrix and visualizations, given 
## the desired design formula. 
vd <- VisualizeDesign(sampleData = meta_noninf_spc, 
                      designFormula = ~ sex)
vd$designmatrix
vd$plotlist

## Note that we can also generate the design matrix like this
model.matrix(~ sex, data = meta_noninf_spc)
```

:::::::::::::::::::::::::::::::::::::::  challenge

### Challenge

With this design, what is the interpretation of the `sexMale` coefficient?

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::  challenge

### Challenge

Set up the design formula to compare male and female spinal cord samples from Day0 as above, but instruct R to not include an intercept in the model. How does this change the interpretation of the coefficients? What contrast would have to be specified to compare the mean expression of a gene between male and female mice? 

:::::::::::::::  solution

### Solution

```{r, echo=TRUE}
meta_noninf_spc <- meta %>% filter(time == "Day0" & 
                                       tissue == "Spinalcord")
meta_noninf_spc
vd <- VisualizeDesign(sampleData = meta_noninf_spc, 
                      designFormula = ~ 0 + sex)
vd$designmatrix
vd$plotlist
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::  challenge

### Challenge

Set up the design formula to compare the three time points (Day0, Day4, Day8) in the male spinal cord samples, and visualize it using `ExploreModelMatrix`.

:::::::::::::::  solution

### Solution

```{r, echo=TRUE}
meta_male_spc <- meta %>% filter(sex == "Male" & tissue == "Spinalcord")
meta_male_spc
vd <- VisualizeDesign(sampleData = meta_male_spc, designFormula = ~ time)
vd$designmatrix
vd$plotlist
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Factorial design without interactions

Next, we again consider only non-infected mice, but fit a model incorporating both sex and tissue as predictors. 
We assume that the tissue differences are the same for both male and female mice, and consequently fit an additive model, without interaction terms. 

```{r}
meta_noninf <- meta %>% filter(time == "Day0")
meta_noninf
vd <- VisualizeDesign(sampleData = meta_noninf, 
                      designFormula = ~ sex + tissue)
vd$designmatrix
vd$plotlist
```

## Factorial design with interactions

In the previous model, we assumed that the tissue differences were the same for both male and female mice. 
To allow for the estimation of sex-specific tissue differences (at the expense of having one additional coefficient to estimate from the data), we can include an interaction term in the model. 

```{r}
meta_noninf <- meta %>% filter(time == "Day0")
meta_noninf

## Define a design including an interaction term
## Note that ~ sex * tissue is equivalent to 
## ~ sex + tissue + sex:tissue
vd <- VisualizeDesign(sampleData = meta_noninf, 
                      designFormula = ~ sex * tissue)
vd$designmatrix
vd$plotlist
```

## Paired design

In this particular data set the samples are paired - the same mice have contributed both the cerebellum and spinal cord samples. 
This information was not included in the previous models. 
However, accounting for it can increase power to detect tissue differences by eliminating variability in baseline expression levels between mice. 
Here, we define a paired design for the female non-infected mice, aimed at testing for differences between tissues after accounting for baseline differences between mice.

```{r}
meta_fem_day0 <- meta %>% filter(sex == "Female" & 
                                     time == "Day0")

# ensure that mouse is treated as a categorical variable
meta_fem_day0$mouse <- factor(meta_fem_day0$mouse)

meta_fem_day0

vd <- VisualizeDesign(sampleData = meta_fem_day0,
                      designFormula = ~ mouse + tissue)
vd$designmatrix
vd$plotlist
```

## Within- and between-subject comparisons

In some situations, we need to combine the types of models considered above. 
For example, let's say that we want to investigate if the tissue differences are different for infected and non-infected female mice. 
In this case, each mice only contributes to one of the infection groups (each mice is either infected or non-infected), but contributes both a cerebellum and a spinal cord sample.
One way to view this type of design is as two paired experiments, one for each infection group (see the [edgeR user guide section 3.5](https://www.bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf)).
We can then easily compare the two tissues in each infection group, and contrast the tissue differences between the infection groups. 

```{r}
meta_fem_day04 <- meta %>% 
    filter(sex == "Female" & 
               time %in% c("Day0", "Day4")) %>%
    droplevels()
# ensure that mouse is treated as a categorical variable
meta_fem_day04$mouse <- factor(meta_fem_day04$mouse)

meta_fem_day04
design <- model.matrix(~ mouse, data = meta_fem_day04)
design <- cbind(design, 
                Spc.Day0 = meta_fem_day04$tissue == "Spinalcord" & 
                    meta_fem_day04$time == "Day0",
                Spc.Day4 = meta_fem_day04$tissue == "Spinalcord" & 
                    meta_fem_day04$time == "Day4")
rownames(design) <- rownames(meta_fem_day04)
design
vd <- VisualizeDesign(sampleData = meta_fem_day04 %>%
                          select(time, tissue, mouse),
                      designFormula = NULL, 
                      designMatrix = design, flipCoordFitted = FALSE)
vd$designmatrix
vd$plotlist
```

:::::::::::::::::::::::::::::::::::::::: keypoints

- The formula framework in R allows creation of design matrices, which details the variables expected to be associated with systematic differences in gene expression levels. 
- Comparisons of interest can be defined using contrasts, which are linear combinations of the model coefficients. 


::::::::::::::::::::::::::::::::::::::::::::::::::


